Designing Types for R, Empirically

### 2020-05-16 paper/main.tex
Fix class name formatting

We breakdown the failed assertions by type in
Table~\ref{tbl:fail-breakdown-0}.  
Accounting for 27.37\% of assertion failures, the most common vilation arises when a value with \CLASS{matrix} is passed where a \VEC\dbl is expected.
Considering these types, we might imagine them to be compatible, however not allowing this coercion was a deliberate design decision:
coercion from vectors to matrices is ad hoc at best, and surely not a
practice codified in the language.  In a similar vein, another popular
failing assertion is checking if a ${\bf dbl}[]$ has type ${\bf int}[]$,
another case of commonly performed coercion.  We did not include these types
of coercions in our type annotation framework as programmers cannot rely on
them, and it is not always the case that the coercions are safe to perform.

@@

\begin{tabular}{llrrr}
\toprule
Passed & Arg Type & Occurrences & \% Total & Cumulative \%\\
\midrule
\VEC\dbl & \CLASS{matrix} & 371,896 & 27.37 & 27.37\\
\VEC\dbl & \VEC\int & 117,212 & 8.62 & 35.99\\
\chr & \CLASS{bignum} | \VEC\raw & 100,100 & 7.37 & 43.36\\
\dbl & \int | \null & 60,578 & 4.46 & 47.81\\
\VEC\dbl & \CLASS{timeSeries} & 58,872 & 4.33 & 52.15\\
\addlinespace
\CLASS{matrix} & \CLASS{timeSeries} & 55,005 & 4.05 & 56.19\\
\VEC\dbl & \dbl & 53,338 & 3.92 & 60.12\\
\dbl & \CLASS{data.frame} & 32,553 & 2.40 & 62.51\\
\VEC\dbl & \CLASS{data.frame} & 31,576 & 2.32 & 64.84\\
\dbl & \int & 20,586 & 1.51 & 66.35\\
\bottomrule
\end{tabular}
% END Autogenerated
\label{tbl:fail-breakdown-0}
\caption{Top types of contract assertion failures.}
\end{table}

@@

Turning our attention now to arguments, we found that only
\PROPARGSFAILINGASSERTS of function parameters exhibited a failed assertion.
Table~\ref{tbl:fail-breakdown-0} showed the raw results from counting
runtime occurrences, but that data alone does not tell the full story, as
some failures may be overrepresented if e.g. a failing contract assertion
was in a loop.  We were interested in knowing for each of the most common
violations in Table~\ref{tbl:fail-breakdown-0}, how many different arguments
had that type, and how many of those exhibited the contract failure in
question.  Table~\ref{tbl:fail-breakdown-1} breaks down the failed assertions by type,
folding away multiple identical failed contract assertions for the same parameter
position.  The first row of this table reads: a value of type
\VEC\dbl was passed to 15 different function parameters expecting a
\CLASS{matrix}, of which there are 1394 in total: 1.08\% of \CLASS{matrix}-typed parameters
were passed \VEC\dbl values.

We see that even though the double vector and matrix issue was wildly
prevalent in the raw, dynamic contract evaluation numbers, the number of
actual function argument types that were violated is very small.  The story
is similar with the double and integer coercion we mentioned earlier: it
represents many dynamic contract failures, but very few of the \VEC\int-typed 
arguments have their contracts violated by \VEC\dbl-typed
values.  Rows 5 and 6 are interesting: we see that rather often arguments
expecting \CLASS{timeSeries} data are passed \VEC\dbl or \CLASS{matrix} values.  This
is a quirk of the \code{timeSeries} package, whose functions often accept
matrices and vectors, converting them to time series in an ad hoc manner.
Note that the code coverage of the \code{timeSeries} tests, examples, and
vignettes on the \code{timeSeries} package code is only 58.24\%, which is
one possible explanation of why these contract failures are occurring: the
types that \typetracer generates are only as good as the test code its run
on.

@@

\begin{tabular}{llrrr}
\toprule
Passed & Arg Type & \# Args Failed & \# Args with Type & \% Failure\\
\midrule
\VEC\dbl & \CLASS{matrix} & 15 & 1394 & 1.08\\
\VEC\dbl & \VEC\int & 44 & 657 & 6.70\\
\chr & \CLASS{bignum} | \VEC\raw & 1 & 3 & 33.33\\
\dbl & \int | \null & 5 & 20 & 25.00\\
\VEC\dbl & \CLASS{timeSeries} &19 & 35 & 54.29\\
\addlinespace
\CLASS{matrix} & \CLASS{timeSeries} &  19 & 35 & 54.29\\
\VEC\dbl & \dbl & 100 & 5021 & 1.99\\
\dbl & \CLASS{data.frame} &  1 & 900 & 0.11\\
\VEC\dbl & \CLASS{data.frame} &  6 & 900 & 0.67\\
\dbl & \int & 48 & 529 & 9.07\\
\bottomrule
\end{tabular}

@@

Table~\ref{tbl:fail-breakdown-2} presents data on the most frequently
violated contracts amongst the most frequently occurring argument types.  We
selected argument types which were in the 90th percentile of argument type
occurrences, computed the most frequent type signature violations among
them, and reported the most frequently violated contracts together with the
type of the value that violated that contract.  The first row of the table
reads: 44 function arguments with \VEC\int type are passed \VEC\dbl  instead, and 657 arguments have \VEC\int type.  Had we failed
to capture some key usage pattern of R with our type annotation framework,
we would likely see it here, and we can see this in action if we consider
Table~\ref{tbl:fail-breakdown-3}, which was obtained identically to
Table~\ref{tbl:fail-breakdown-2} with a percentile of 80th percentile
instead.  The most frequent argument type violation pattern in that of \CLASS{data.frame} values
 passed to arguments expecting tibbles, which are an extension of \CLASS{data.frame} built in the \code{tidyverse} package suite: 
 this occurs in only 22\%
of such arguments, and represents cases where tests did not adequately cover
all valid function inputs.  Separate from the issue of testing, we can
capture this behaviour with user-defined subtyping, as data frames are
subtypes of tibbles, but this necessitates a proper object-oriented type
system for R which is outside the scope of this work.

@@

\begin{tabular}{llrrr}
\toprule
Passed & Arg Type & \# Args Failed & \# Args with Type & \% Failure\\
\midrule
\CLASS{data.frame} & \CLASS{data.frame, tbl, tbl\_df} & 63 & 284 & 22.18\\
\CLASS{matrix} & \CLASS{array} & 13 & 70 & 18.57\\
\NAVEC\lgl & \lgl | \null & 9 & 49 & 18.37\\
\VEC\chr & \LIST{\VEC\chr} & 20 & 109 & 18.35\\
\VEC\clx & \clx & 7 & 42 & 16.67\\
\bottomrule
\end{tabular}
% END Autogenerated
 \label{tbl:fail-breakdown-3}
 \caption{What is the highest failure rate among popular argument types? For argument signatures whose frequency is in the 80th percentile.}
 \end{table}

>>>>

We breakdown the failed assertions by type in
Table~\ref{tbl:fail-breakdown-0}.  
Accounting for 27.37\% of assertion failures, the most common vilation arises when a value with \CLASS{\texttt{matrix}} is passed where a \VEC\dbl is expected.
Considering these types, we might imagine them to be compatible, however not allowing this coercion was a deliberate design decision:
coercion from vectors to matrices is ad hoc at best, and surely not a
practice codified in the language.  In a similar vein, another popular
failing assertion is checking if a ${\bf dbl}[]$ has type ${\bf int}[]$,
another case of commonly performed coercion.  We did not include these types
of coercions in our type annotation framework as programmers cannot rely on
them, and it is not always the case that the coercions are safe to perform.

@@

\begin{tabular}{llrrr}
\toprule
Passed & Arg Type & Occurrences & \% Total & Cumulative \%\\
\midrule
\VEC\dbl & \CLASS{\texttt{matrix}} & 371,896 & 27.37 & 27.37\\
\VEC\dbl & \VEC\int & 117,212 & 8.62 & 35.99\\
\chr & \CLASS{\texttt{bignum}} | \VEC\raw & 100,100 & 7.37 & 43.36\\
\dbl & \int | \null & 60,578 & 4.46 & 47.81\\
\VEC\dbl & \CLASS{\texttt{timeSeries}} & 58,872 & 4.33 & 52.15\\
\addlinespace
\CLASS{\texttt{matrix}} & \CLASS{\texttt{timeSeries}} & 55,005 & 4.05 & 56.19\\
\VEC\dbl & \dbl & 53,338 & 3.92 & 60.12\\
\dbl & \CLASS{\texttt{data.frame}} & 32,553 & 2.40 & 62.51\\
\VEC\dbl & \CLASS{\texttt{data.frame}} & 31,576 & 2.32 & 64.84\\
\dbl & \int & 20,586 & 1.51 & 66.35\\
\bottomrule
\end{tabular}
% END Autogenerated
\label{tbl:fail-breakdown-0}
\caption{Top types of contract assertion failures.}
\end{table}

@@

Turning our attention now to arguments, we found that only
\PROPARGSFAILINGASSERTS of function parameters exhibited a failed assertion.
Table~\ref{tbl:fail-breakdown-0} showed the raw results from counting
runtime occurrences, but that data alone does not tell the full story, as
some failures may be overrepresented if e.g. a failing contract assertion
was in a loop.  We were interested in knowing for each of the most common
violations in Table~\ref{tbl:fail-breakdown-0}, how many different arguments
had that type, and how many of those exhibited the contract failure in
question.  Table~\ref{tbl:fail-breakdown-1} breaks down the failed assertions by type,
folding away multiple identical failed contract assertions for the same parameter
position.  The first row of this table reads: a value of type
\VEC\dbl was passed to 15 different function parameters expecting a
\CLASS{\texttt{matrix}}, of which there are 1394 in total: 1.08\% of \CLASS{\texttt{matrix}}-typed parameters
were passed \VEC\dbl values.

We see that even though the double vector and matrix issue was wildly
prevalent in the raw, dynamic contract evaluation numbers, the number of
actual function argument types that were violated is very small.  The story
is similar with the double and integer coercion we mentioned earlier: it
represents many dynamic contract failures, but very few of the \VEC\int-typed 
arguments have their contracts violated by \VEC\dbl-typed
values.  Rows 5 and 6 are interesting: we see that rather often arguments
expecting \CLASS{\texttt{timeSeries}} data are passed \VEC\dbl or \CLASS{\texttt{matrix}} values.  This
is a quirk of the \code{timeSeries} package, whose functions often accept
matrices and vectors, converting them to time series in an ad hoc manner.
Note that the code coverage of the \code{timeSeries} tests, examples, and
vignettes on the \code{timeSeries} package code is only 58.24\%, which is
one possible explanation of why these contract failures are occurring: the
types that \typetracer generates are only as good as the test code its run
on.

@@

\begin{tabular}{llrrr}
\toprule
Passed & Arg Type & \# Args Failed & \# Args with Type & \% Failure\\
\midrule
\VEC\dbl & \CLASS{\texttt{matrix}} & 15 & 1394 & 1.08\\
\VEC\dbl & \VEC\int & 44 & 657 & 6.70\\
\chr & \CLASS{\texttt{bignum}} | \VEC\raw & 1 & 3 & 33.33\\
\dbl & \int | \null & 5 & 20 & 25.00\\
\VEC\dbl & \CLASS{\texttt{timeSeries}} &19 & 35 & 54.29\\
\addlinespace
\CLASS{\texttt{matrix}} & \CLASS{\texttt{timeSeries}} &  19 & 35 & 54.29\\
\VEC\dbl & \dbl & 100 & 5021 & 1.99\\
\dbl & \CLASS{\texttt{data.frame}} &  1 & 900 & 0.11\\
\VEC\dbl & \CLASS{\texttt{data.frame}} &  6 & 900 & 0.67\\
\dbl & \int & 48 & 529 & 9.07\\
\bottomrule
\end{tabular}

@@

Table~\ref{tbl:fail-breakdown-2} presents data on the most frequently
violated contracts amongst the most frequently occurring argument types.  We
selected argument types which were in the 90th percentile of argument type
occurrences, computed the most frequent type signature violations among
them, and reported the most frequently violated contracts together with the
type of the value that violated that contract.  The first row of the table
reads: 44 function arguments with \VEC\int type are passed \VEC\dbl  instead, and 657 arguments have \VEC\int type.  Had we failed
to capture some key usage pattern of R with our type annotation framework,
we would likely see it here, and we can see this in action if we consider
Table~\ref{tbl:fail-breakdown-3}, which was obtained identically to
Table~\ref{tbl:fail-breakdown-2} with a percentile of 80th percentile
instead.  The most frequent argument type violation pattern in that of \CLASS{\texttt{data.frame}} values
passed to arguments expecting tibbles, which are an extension of \CLASS{\texttt{data.frame}} built in the \code{tidyverse} package suite: 
 this occurs in only 22\%
of such arguments, and represents cases where tests did not adequately cover
all valid function inputs.  Separate from the issue of testing, we can
capture this behaviour with user-defined subtyping, as data frames are
subtypes of tibbles, but this necessitates a proper object-oriented type
system for R which is outside the scope of this work.

@@

\begin{tabular}{llrrr}
\toprule
Passed & Arg Type & \# Args Failed & \# Args with Type & \% Failure\\
\midrule
\CLASS{\texttt{data.frame}} & \CLASS{\texttt{data.frame}, \texttt{tbl}, \texttt{tbl\_df}} & 63 & 284 & 22.18\\
\CLASS{\texttt{matrix}} & \CLASS{\texttt{array}} & 13 & 70 & 18.57\\
\NAVEC\lgl & \lgl | \null & 9 & 49 & 18.37\\
\VEC\chr & \LIST{\VEC\chr} & 20 & 109 & 18.35\\
\VEC\clx & \clx & 7 & 42 & 16.67\\
\bottomrule
\end{tabular}
% END Autogenerated
 \label{tbl:fail-breakdown-3}
 \caption{What is the highest failure rate among popular argument types? For argument signatures whose frequency is in the 80th percentile.}
 \end{table}
